Settler colonial projects explain variation in US state formation in the early 20th c.

### 2024-03-06 ms.Rmd
random updates that I don't remember

---
title: "Settler colonial projects explain variation in US state formation in the early 20th c."
author: "Frank Edwards"
date: "last updated `r Sys.Date()`"
output: bookdown::html_document2
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)
library(tidybayes)
library(rstan)

options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r read, include = F, message = F}
### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

# read data from make_* scripts
pop<-read_csv("./data/temp_population_ts.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")
bs<-read_csv("./data/temp_bs_ts.csv")
agents<-read_csv("./data/temp_agents.csv")

## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents)

dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents)
```

```{r run-models, cache = T, include = F, message = F}
#source("models.R")
```

@@

Under sharecropping regimes in the US South, white settler elites sought to create and reproduce brutal and paternalistic regimes of informal and formal social regulation that kept Black workers and their families dependent on and subordinate to white interests. Industrial and social white elites in the North sought to discipline large numbers of European immigrants toward nascent ideals of a white American settler polity oriented toward a narrow gendered division of labor and subordination to the interests of industrial capital. In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers. White settlers and settler corporate interests flooded recently disposessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.

@@

Consistent with the historigraphy on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure. State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects. In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding shcools. In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies. Social control and reproduction instead relied on a coordinated formal and informal system of racial terror that developed alongside a relatively weak set of state social policy institutions.

@@

We focus our attention on three interconnected settler colonial structures and processes that had large impacts on social life in the early 20th century US: 1) the active dispossession and erasure of Native nations and the development of settler enterprises and communities; 2) the exploitation of Black labor for agricultural production and the fortification of an apartheid social order; and 3) the coercive and paternalistic assimilation of recent immigrants from Eastern and Southern Europe into an urban industrial settler colonial polity. We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation. Or, put more directly, that THE LOCAL SALIENCE OF PARTICULAR CONSITUENTS / STRUCTURES IS RELATED TO STATE FORMATION.

These are three of the projects that were critical to settler colonialism in the early 20th c US. Not the only ones of course! Colonial processes deeply impacted the experiences of a broad range of groups and produced a range of distinct racialized social structures (CITES FOR MEXICO BORDER, MIGRANT WORKERS, CHINESE EXCLUSION, JAPANESE INTERNMENT, PHILLIPINES COLONIZATION, AND OTHERS). *Here, we narrowly focus on three settler colonial projects to illustrate how a macro-comparative approach can be helpful in validating core findings in historical scholarship and in incorporating the insights of critical scholarship on settler colonization and racial capitalism into some of the core methods of political, economic, and comparative historical sociology REVISE.*

@@

The geographic distribution of the local implementation of the three focal settler projects we focus on is crudely illustrated in Figure \@ref(fig:three-maps). The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast. While Native land disposession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries. In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862.

The Jim Crow system of apartheid was developed and becoming deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people. As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.

And the Industrial Revolution, coupled with a wave of crises across Eastern and Southern Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becomingly increasingly urban and foreign born.

@@

```{r three-maps, fig.cap = "Geographic and demographic contexts in 1910 for settler colonial governance"}
### ADD DISPOSESSION
p1 <- dist %>% 
  select(STUSPS, year,
         morill_acres_cum, homestead_acres_cum,
         total_acres_2021) %>% 
  filter(year == 1910) %>% 
  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
  left_join(st) %>% 
  ggplot(aes(fill = dist / 1e6,
             geometry = geometry)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Acres\n(millions)",
       subtitle = "1. Lands patented through\nMorill and Homestead acts")

p2 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = blk / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "2. Black population")

p3 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = urban_immig_pop / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "3. Urban immigrant population")

grid.arrange(p1, p2, p3, nrow = 1)
```

## Settling the frontier

OVERVIEW: BUILDING THE SETTLER NATION STATE, ERASING NATIVE NATIONS

Dispossession

Redistribution

THE MORILL, HOMESTEAD AND RR ACTS

Nation building

Forced assimilation

```{r fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))

st_acres<-data.frame(state = st$STUSPS,
                     total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))

### royce map data
## available here: https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
## and archived here: https://data.nativeland.info/dataset/indian-land-cessions-in-the-united-states-1784-1894/resource/e2b874d5-3c39-4287-bf89-d4ef14b87c83
cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")

cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
  mutate(time = as_date(TribalCededLandsTableCessDate1),
         year = year(time)) %>% 
  rename(CESSNUM = TribalCededLandsNewCESSNUM) %>% 
  select(CESSNUM, year)

cessions<-cessions %>% 
  left_join(cessions_dat)

cessions<-cessions %>% 
  st_transform(8528)

cessions_diff<-st_difference(st, st_union(cessions))

cessions_diff<-cessions_diff %>% 
  mutate(year = NA)

### layer state boundaries
cessions<-st_intersection(cessions, st)
### add in settler claimed territories as of 1796
cessions2 <- cessions %>%  
  bind_rows(cessions_diff %>%  
              mutate(year = NA))

ggplot(cessions2,
       aes(fill = year)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "Year") 
```

MANY OF THE LOCATIONS THESE FIRST GENERATION SETTLERS WERE GOING TO WERE RECENT DISPOSESSIONS, PARTICULARLY CIVIL WAR ERA. IN 1900, MANY OF THESE LANDS HAD BEEN CLAIMED BY US IN VERY RECENT PAST, WITHIN 50 YEARS. SECOND / THIRD WAVE OF WESTERN SETTLERS. BUILDING THE SETTLER COLONIAL POLITY / NATION STATE AS KEY AGENDAS.

```{r fig.cap = "Disposession of Native lands in the 19th c."}
p1<-ggplot(disp,
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = cumulative_ceded / 1e6)) + 
  labs(x = "Year",
       y = "Acres (millions)") + 
  lims(y = c(0, 150)) + 
  facet_wrap(region~division) + 
  labs(subtitle = "1. Total area")

p2<-ggplot(disp,
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = prop_ceded)) + 
  labs(x = "Year",
       y = "Proportion") + 
  facet_wrap(region~division) + 
  labs(subtitle = "2. Proportion of state area")

grid.arrange(p1, p2, ncol = 2)
```

This shows the timing and spatial distribution in cumulative land disposession by state and region. Panel 1 shows the lands claimed by the US in millions of acres by state from 1796 - 1894. Panel 2 shows these patterns of Native land loss as a proportion of the contemporary areas of US states.

Clear regional patterns here. Land loss occurred earlier in states closer to the Atlantic coast. Dispossession accelerated in the North Central and Western regions throughout the middle portion of the 19th c.

```{r fig.cap = "Lands patented under two legislative mechanisms for the distribution of disposessed Native Lands to settlers, cuulative total 1930"}
pdat<-disp %>% 
  select(STUSPS, total_acres_2021) %>% 
  distinct() %>% 
  left_join(dist %>% 
              filter(year == 1930) %>% 
              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
  left_join(st)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = morill_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "1. Morill Act") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = homestead_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "2. Homestead Act") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, ncol = 2)
```

## Entrenching the exploitation of Black labor and marginalization of Black social and political power

Between 1900 and 1930, settler elites in the the US South were busy entrenching Jim Crow and profiting from the deeply exploitative sharecropping system of agricultural production.

Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period. In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agrictultural production. In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States.

### The sharecropping economy

DEFINE SYSTEM

SUMMARIZE LITERATURE

### The white southern polity

SUMMARIZE LIT ON SOUTHERN WHITENESS AND COHERENCE OF WHITE SETTLER POLITY IN US SOUTH

### The early and mid-career of Jim Crow

Racial order, law, and terrorism

ESP THINK ON BUREAUCRACY, INFORMALITY OF SOCIAL CONTROL, COLLABORATION BETWEEN FORMAL AND INFORMAL, VIGILANTISM AND LAW

EDUCATION AND JIM CROW

SOCIAL WORK AND JIM CROW

```{r jim-crow-maps, fig.cap = "Black tenant farming and immigrant status of white settlers, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = blk_sharecrop / 1e5)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(hundred thousands)",
       subtitle = "1. Black population living on tenant farms") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / (white_immig + white_non_immig))) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Proportion",
        subtitle = "2. Proportion of white poulation foreign born") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, nrow = 1)

@@

## Incorporating urban white immigrants into the settler polity and industrial working class

@@

```{r immig-maps, fig.cap = "White immigrant population and urban population, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)

p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
       subtitle = "1. White, foreign-born") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = urban_pop / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
        subtitle = "2. Urban") + 
  theme(legend.position = "bottom")

>>>>

---
title: "Settler colonial projects explain variation in US state formation in the early 20th c."
author: "Frank Edwards and Theresa Rocha Beardall"
date: "last updated `r Sys.Date()`"
output: bookdown::pdf_document2
---
```{r setup, include = F, message = F, cache = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
library(brms)
library(tidycensus)
library(tigris)
library(units)
library(sf)
library(tidybayes)
library(rstan)
library(knitr)
library(trackdown)
# trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
# trackdown::update_file(file = "ms.Rmd", 
#                        hide_code = T,
#                        rich_text = T)
options(tigris_use_cache = T)
options(scipen = 999)
options(dplyr.print_max = 30)
map<-map_data("state")
theme_set(theme_minimal())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
# set hook for thousands comma
knit_hooks$set(inline = function(x) {
prettyNum(round(x,2), big.mark=",")
})
```

```{r read, include = F, message = F}
### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))
# read data from make_* scripts
pop<-read_csv("./data/temp_population_ts.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")
bs<-read_csv("./data/temp_bs_ts.csv")
agents<-read_csv("./data/temp_agents.csv")
## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents %>% 
              select(-state))
dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents %>% 
              select(-state))
```

```{r run-models, cache = T, include = F, message = F}
source("models.R")
```

@@

Under sharecropping regimes in the US South, white settler elites sought to create and reproduce brutal and paternalistic regimes of informal and formal social regulation that kept Black workers and their families dependent on and subordinate to white interests. Industrial and social white elites in the North sought to discipline large numbers of European immigrants toward nascent ideals of a white American settler polity oriented toward a narrow gendered division of labor and subordination to the interests of industrial capital. In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers. White settlers and settler corporate interests flooded recently dispossessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.

@@

@@

We focus our attention on three interconnected settler colonial structures and processes that had large impacts on social life in the early 20th century US: 1) the active dispossession and erasure of Native nations and the development of settler enterprises and communities; 2) the exploitation of Black labor for agricultural production and the fortification of an apartheid social order; and 3) the coercive and paternalistic assimilation of recent immigrants from Eastern and Southern Europe into an urban industrial settler colonial polity. We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation. In more general terms, we argue that the local salience and configuration of key social structures, like settler colonization, are closely related to local trajectories for state formation and institution building.

Consistent with the historiography on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure. State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects. Our work bridges these sometimes disparate historical accounts by showing that the local interests of settler elites were closely tied to the kinds of social and racial projects that the early carceral and welfare states engaged in.

@@

The geographic distribution of the local implementation of the three focal settler projects we focus on is crudely illustrated in Figure \@ref(fig:three-maps). The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast. While Native land dispossession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries. In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862.

```{r three-maps, fig.cap = "Three focal contexts for settler colonial governance, 1910"}
### ADD DISPOSESSION
p1 <- dist %>% 
  select(STUSPS, year,
         morill_acres_cum, homestead_acres_cum,
         total_acres_2021) %>% 
  filter(year == 1910) %>% 
  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
  left_join(st) %>% 
  ggplot(aes(fill = dist / 1e6,
             geometry = geometry)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Acres\n(millions)",
       subtitle = "1. Lands patented through\nMorill and Homestead acts")
p2 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = blk / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "2. Black population")
p3 <- st %>% 
  left_join(pop %>% 
              filter(year == 1910)) %>% 
  ggplot(aes(fill = urban_immig_pop / tot_pop)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Proportion",
       subtitle = "3. Urban immigrant population")
grid.arrange(p1, p2, p3, nrow = 1)
```

The Jim Crow system of apartheid was developed and became deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people. As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.

And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becoming increasingly urban and foreign born.

@@

## Building the settler nation-state

KEY CHARARCTERISTICS: RECENT OR ONGOING DISPOSESSION / REMOVAL, RECENTLY ADMITTED STATES, RELATIVELY SPARSE WHITE POPS. ELIMINATION / REPLACEMENT AS KEY DYNAMIC.

In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding schools.

```{r cessions-map, fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))
st_acres<-data.frame(state = st$STUSPS,
                     total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))
### royce map data
## available here: https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
## and archived here: https://data.nativeland.info/dataset/indian-land-cessions-in-the-united-states-1784-1894/resource/e2b874d5-3c39-4287-bf89-d4ef14b87c83
cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")
cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
  mutate(time = as_date(TribalCededLandsTableCessDate1),
         year = year(time)) %>% 
  rename(CESSNUM = TribalCededLandsNewCESSNUM) %>% 
  select(CESSNUM, year)
cessions<-cessions %>% 
  left_join(cessions_dat)
cessions<-cessions %>% 
  st_transform(8528)
cessions_diff<-st_difference(st, st_union(cessions))
cessions_diff<-cessions_diff %>% 
  mutate(year = NA)
### layer state boundaries
cessions<-st_intersection(cessions, st)
### add in settler claimed territories as of 1796
cessions2 <- cessions %>%  
  bind_rows(cessions_diff %>%  
              mutate(year = NA))
ggplot(cessions2,
       aes(fill = year)) + 
  geom_sf() + 
  theme_void() + 
  scale_fill_viridis_c() + 
  labs(fill = "Year")

@@

Figures \@ref(fig:cessions-map) and \@ref(fig:cessions-ts) show the temporal and spatial process of the dispossession of lands from Native Nations by the US Government between 1795 and 1893 (DATA FROM USFS VIA Lib Of Cong). These maps show a clear difference in the timing of land cessions as a function of place. By 1795, the US Government claimed nearly all lands East of the Appalachian Mountains, including large parts of contemporary Kentucky and Tennessee. Land dispossession throughout the South and the Great Lakes regions accelerated quickly during the early 19th century. Throughout the middle of the 19th century, there were large-scale efforts by the federal government and vigilante settlers to dispossess tribal lands in California, the Midwest, Great Plains, and Mountain West. By 1895, nearly all lands in the contemporary lower 48 US states had been claimed by the federal government as US territory.

```{r cessions-ts, fig.cap = "Cumulative disposession of Native lands in the 19th c. by US region and state as proportion of contemporary state area. Each line represents one state."}
# p1<-ggplot(disp,
#        aes(x = year,
#            group = STUSPS)) + 
#   geom_line(aes(y = cumulative_ceded / 1e6)) + 
#   labs(x = "Year",
#        y = "Acres (millions)") + 
#   lims(y = c(0, 150)) + 
#   facet_wrap(region~division) + 
#   labs(subtitle = "1. Total area")
p2<-ggplot(disp %>% 
             filter(year > 1795),
       aes(x = year,
           group = STUSPS)) + 
  geom_line(aes(y = prop_ceded)) + 
  labs(x = "Year",
       y = "Proportion of state area") + 
  facet_wrap(region~division) 
p2
```

The Federal government passed a series of laws in 1862 designed to rapidly redistribute dispossessed Native lands to white settlers and to large corporate and state institutions. The Homestead, Morill, and Pacific Railway Acts of 1862 provided legal frameworks through which expropriated Native lands could be rapidly redistributed to white settlers, large corporations, and Universities. Figure \@ref(fig:hs-morill-map) shows the total number of acres within a state that had been granted to settlers and settler institutions by 1930. More than 10 million acres of Native lands had been redistributed to US land grant Universities at little or no cost for their use or liquidation by 1930 under the Morill act. Under the Homestead Act, more than 255 million acres of Native land had been transferred to settlers by 1930. States in the Midwest, Great Plains, Mountain West, and West Coast saw the highest volumes of land redistribution under these laws.

```{r hs-morill-map, fig.cap = "Lands patented under two legislative mechanisms for the redistribution of disposessed Native lands to settlers, cumulative total 1862 - 1930"}
pdat<-disp %>% 
  select(STUSPS, total_acres_2021) %>% 
  distinct() %>% 
  left_join(dist %>% 
              filter(year == 1930) %>% 
              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
  left_join(st)
p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = morill_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "1. Morill Act") + 
  theme(legend.position = "bottom")

p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = homestead_acres_cum / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Acres (millions)",
       subtitle = "2. Homestead Act") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, ncol = 2)
```

## Jim Crow

KEY CHARACTERISTICS: ANTI-BLACK SOCIAL, LEGAL, ECONOMIC REGIME. WHITE SETTLER HOMOGENEITY

By 1877 efforts to build effective racial democracy through post-Civil War Reconstruction were abandoned by the federal government. By the turn of the century, white elites had legislated and entrenched a regime of brutal racial exploitation and apartheid across the US South (DU BOIS, FONER, FRANKLIN, CURTIN).

DEFINE JIM CROW REGIME. APARTHEID, FORMAL / INFORMAL TERROR REGIME, BLACK LABOR EXPLOITATION, PRIMARILY AG through SHARECROPPING.

READ APD CITATIONS FROM ANN REVIEWS ON JIM CROW, APD, RACISM. DOES APD SUPPORT THE WEAK STATE THESIS?

In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies. Social control and reproduction instead relied on a coordinated formal and informal system of racial terrorism and apartheid that developed alongside a relatively weak set of state social policy institutions (QUADAGNO, KATZNELSON).

Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period. In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agricultural production. In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States.

```{r jim-crow-maps, fig.cap = "Black tenant farming and immigrant status of white settlers, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)
p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = blk_sharecrop / 1e5)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(hundred thousands)",
       subtitle = "1. Black population living on tenant farms") + 
  theme(legend.position = "bottom")
p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / (white_immig + white_non_immig))) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Proportion",
        subtitle = "2. Proportion of white population foreign born") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, nrow = 1)
```

## Regulating and incorporating urban white immigrants

KEY CHARACTERISTICS: URBANIZATION, INDUSTRIALIZATION, MORAL PANICS OVER IMMIGRANT CULTURAL DIFFERENCE. (TWO ROUTES, REACTIONARY DISCIPLINE, LIBERAL WELFARE PATERNALISM, BOTH ACTIVE).

In the rapidly urbanizing portions of the Northeast, Midwest, and West Coast, we see heavy investment in what has become the ideal type of the Progressive social state: a robust and professionalizing police force and a broad paternalistic welfare state. These institutions were developed by elites, in part, to regulate urban disorder discursively linked to growing populations of European immigrants and Black migrants from the US South, and with rapidly assimilating these groups into the norms and ideologies of industrial capitalism and the settler nation.

@@

```{r immig-maps, fig.cap = "White immigrant population and urban population, 1910"}
pdat<-st %>% 
  left_join(pop) %>% 
  filter(year == 1910)
p1<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = white_immig / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
       subtitle = "1. White, foreign-born") + 
  theme(legend.position = "bottom")
p2<-ggplot(pdat) + 
  geom_sf(aes(geometry = geometry,
              fill = urban_pop / 1e6)) + 
  scale_fill_viridis_c() + 
  theme_void() + 
  labs(fill = "Population\n(millions)",
        subtitle = "2. Urban") + 
  theme(legend.position = "bottom")
grid.arrange(p1, p2, nrow = 1)
```

Progressive reformers in the Northeast and Midwest were preoccupied with solving a battery of social problems associated with urban life, industrial capitalism, and their perceptions of the culture and behavior of Southern and Eastern European immigrants.

HIGH LEVEL SUMMARY OF 1st WAVE WHITE FEMINISM AND SOCIAL WORK VIA GORDON, HARTMAN, OTHERS.

The emerging fields of sociology, social work, and criminology also sought to apply scientific principles to the organization and practice of policing and corrections. HIGH LEVEL SUMMARY OF PROG ERA POLICING, JUV COURT, PRISON REFORM. LINK TO GOTTSCHALK ON CARCERAL STATE EMERGENCE FROM THESE MOVEMENTS.

# Patterns in the development of social policy infrastructure 1900 - 1930

We define social state capacity as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a detailed review of theory and measurement). Capacity is a function of both physical infrastructure and human resources available to state actors. We focus on a broad set of social policy systems including education, social welfare, and criminal justice.

We rely on three sets of measures associated with the capacity of states to accomplish social policy goals: 1) employment rates in key street-level social policy positions as a measure of the relative size of bureaucracies and magnitude of human effort available to policy actors, 2) populations housed in institutional group quarters as a measure of the scope of particular kinds of physical infrastructure used by the welfare and carceral systems during the Progressive Era, and 3) counts of operating Indian boarding schools as indicators of physical infrastructure directed at coerced assimilation and erasure of Indigenous peoples. Taken together, these three sets of measures provide a diverse and broad set of insights into spending on the basic infrastructure of social governance during the early 20th century.

## Employment in education, policing, and social welfare: 1900 - 1930

There was dramatic growth in public sector employment in general (CLEAR FROM FED GOVT EMPLOYMENT, OTHER CITATIONS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era. *These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (GENERAL LIT ON GOVT COMPLEXITY AND ARTICULATION 1900 - 1930).*

In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930 in panel 1, and the proportional change in employment rates by category over time in panel 2. Employment of teachers increased from a rate of 5.5 per 1,000 population in 1900 to a rate of 8.6 per 1,000 in 1930, a 56 percent increase. Employment of police doubled during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930. Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900). However, by 1930, the Census recorded about 17,000 social workers nationally, an increase of about 30 fold in terms of employment rates per capita. CHECK THESE NUMBERS
